M-Estimation of Principal Effects using Principal Scores

### 2022-08-03 writeUps/paper.tex
fixing git stuff

In a recent educational field experiment, middle school students completing their math schoolwork on a computer program were randomized between two conditions. In one condition, students were instructed to work on a fixed number of math problems, and later informed of their scores. In a second condition, students received immediate, informative error messages after entering a wrong answer, and were required to enter a correct answer in order to proceed. Students in the second treatment arm also had access to a series of hints for each problem, the last of which contained the correct answer. Students who ``bottomed out'' on a problem, i.e. requested all available hints on a problem, and were told the correct answer, did not receive credit for that problem. Students remained in their randomly-assigned condition over the course of a school year, and were assessed with a 10-item posttest administered within the computer program, as well with state standardized tests.

@@

Analysis models had access to treatment assignment $Z$, outcomes $Y$,
principal stratum $S$ for members of the treatment group, and
covariates.
Even though three covariates, $x_1$, $x_2$ and $x_3$ were included as
predictors of principal stratum $S$ and outcome $Y$ in the data
generating model, both analysis approaches only had access to the
first two covariates, $x_1$ and $x_2$.
Hence, principal ignorability was violated---$S$ and $Y$ were
``confounted'' by unobserved variable $x_3$.

Simulated data were analyzed first with a ``stacked equations''/M-Estimation/GEE
approach, using the \texttt{geex} package in \texttt{R}
\citep{geex,rcite}, and then with a more conventional maximum
likelihood approach via \texttt{rstan} \citep{rstan}.

Both methods assumed (correctly) that there were no interactions
between covariates $x_1$ and $x_2$ and treatment assignment (after
accounting for $S$).

In the GEE approach included a set of nine estimating equations.
First, a GEE logistic regression fit to the treatment group to
estimate principal scores
$\ppi=logit^{-1}(\beta_0+\beta_1x_{1i}+\beta_2x_{2i})$ (three parameters):
\begin{align*}
  \sum_iZ_i\left[S_i-\ppi\right]&=0\\
  \sum_iZ_ix_{1i}\left[S_i-\ppi\right]&=0\\
  \sum_iZ_ix_{2i}\left[S_i-\ppi\right]&=0
\end{align*}
Next, four equations to estimate the outcome regression in the
treatment group, including terms for $S$:
\begin{align*}
  \sum_i
  Z_i\left[Y_i-(\gamma_1x_{1i}+\gamma_2x_{2i}+S_i\mu_{11}+(1-S_i)\mu_{10})\right]&=0\\
  \sum_i
  Z_ix_{1i}\left[Y_i-(\gamma_1x_{1i}+\gamma_2x_{2i}+S_i\mu_{11}+(1-S_i)\mu_{10})\right]&=0\\
  \sum_i
  Z_ix_{2i}\left[Y_i-(\gamma_1x_{1i}+\gamma_2x_{2i}+S_i\mu_{11}+(1-S_i)\mu_{10})\right]&=0\\
  \sum_i
  Z_iS_i\left[Y_i-(\gamma_1x_{1i}+\gamma_2x_{2i}+S_i\mu_{11}+(1-S_i)\mu_{10})\right]&=0
\end{align*}

Finally, two equations to estimate the mixture model in the control
group, for whom $S$ is unobserved:
\begin{align*}
  \sum_i
  (1-Z_i)\left[Y_i-(\gamma_1x_{1i}+\gamma_2x_{2i}+\ppi\mu_{01}+(1-\ppi)\mu_{00})\right]&=0\\
  \sum_i
  (1-Z_i)\left[Y_i-(\gamma_1x_{1i}+\gamma_2x_{2i}+\ppi^2\mu_{01}+(\ppi-\ppi^2)\mu_{00})\right]&=0
\end{align*}
The average treatment effect for $S=0$ was $\mu_{10}-\mu_{00}$ and for
$S=1$ was $\mu_{11}-\mu_{01}$.

Next, we estimated effects with MLE, using the built-in optimizer of
\texttt{rstan}.

In this model we regressed $S$ and $Y$ for the treatment group on an
intercept, $x_1$, and $x_2$ using a logistic and normal linear
specification, respectively.
Then, we specified a two-component mixture model for control
outcomes.

\section{Data Analysis}

@@

\subsection{Proof for Proposition \ref{prop:reg0}}
Carrying out those replacements in \eqref{eq:estEq0}, and rearranging\footnote{Technically, replacing $\tilde{\Psi}_i$ with $\Psi_i=\begin{psmallmatrix} 1 & 0&1&0\\ 0&1&0&1\\0&0&1&0\\0&0&0&1\end{psmallmatrix}\tilde{\Psi}_i$.} gives an equivalent set of estimating equations:
\begin{equation}\label{eq:estEq1}
\sum_{i=1}^{n_C}\Psi_i=\sum_{i=1}^{n_C}\begin{pmatrix}
    Y_i-\muc0-\ri(\muc1-\muc0)-Z_i(\mut0-\muc0)-Z_i\ri(\mut1-\mut0-\muc1+\muc0)\\
    \ri Y_i-\ri\muc0-\ri^2(\muc1-\muc0)-Z_i\ri(\mut0-\muc0)-Z_i\ri^2(\mut0-\mut1-\muc0+\muc1)\\
    Z_iY_i-Z_i\mut0-Z_i\ri (\mut1-\mut0)\\
    Z_i\ri Y_i -Z_i\ri\mut0-Z_i\ri^2(\mut1-\mut0)
  \end{pmatrix}=\begin{pmatrix} 0\\0\\0\\0\end{pmatrix}
\end{equation}

>>>>

@@

\section{Application: Estimating Treatment Effects for Implementation Subgroups}

In a recent educational field experiment, middle school students completing their math schoolwork on a computer program were randomized between two conditions. In one condition (which we arbitrarily label ``treatment''), students were instructed to work on a fixed number of math problems, and later informed of their scores. In a second condition, students received immediate, informative error messages after entering a wrong answer, and were required to enter a correct answer in order to proceed. Students in the second treatment arm (labeled ``control'') also had access to a series of hints for each problem, the last of which contained the correct answer. Students who ``bottomed out'' on a problem, i.e. requested all available hints on a problem, and were told the correct answer, did not receive credit for that problem. Students remained in their randomly-assigned condition over the course of a school year, and were assessed with a 10-item posttest administered within the computer program, as well with state standardized tests.

For the purpose of demonstrating GEEPERS, we gathered dichotomized data on which students in the treatment condition requested at least one bottom-out hint.
That is, $\st=1$ for students who, if assigned to the treatment condition, would request at least one bottom-out hint, and $\st=0$ for students who would not.
The outcome of interest $Y$ in our example is students' total scores on a ten-item posttest completed within the online tutoring system; $Y$ is an integer ranging from 0 to 10.
The goal of our analysis will be to determine the principal effects $\tau_1=\EE[\yt-\yc|\st=0]$ and $\tau_0=\EE[\yt-\yc|\st=0]$, the average effects of assignment to the treatment condition for bottom-outers and non-bottom-outers.
We also gathered data on baseline covariates: total scores on a 10-item pretest, fifth-grade stanardized test scores, indicator variables for if students have an early intervention program (EIP)

or individualized educational plan (IEP) in place, if students are learning English as a second language (ESOL)students'

empress
Mperes

M-estimation of Principal

peseer per

M estimation regression principal effects

M

@@

\subsection{Proof for Proposition \ref{prop:reg0}}
Carrying out those replacements in \eqref{eq:estEq0}, and rearranging\footnote{Technically, replacing $\tilde{\Psi}_i$ with $\Psi_i=\begin{psmallmatrix} 1 & 0&1&0\\ 0&1&0&1\\0&0&1&0\\0&0&0&1\end{psmallmatrix}\tilde{\Psi}_i$.} gives an equivalent set of estimating equations:
\begin{equation}\label{eq:estEq1}
\sum_{i=1}^{n_C}\Psi_i=\sum_{i=1}^{n_C}\begin{pmatrix}
    Y_i-\muc0-\ri(\muc1-\muc0)-Z_i(\mut0-\muc0)-Z_i\ri(\mut1-\mut0-\muc1+\muc0)\\
    \ri Y_i-\ri\muc0-\ri^2(\muc1-\muc0)-Z_i\ri(\mut0-\muc0)-Z_i\ri^2(\mut0-\mut1-\muc0+\muc1)\\
    Z_iY_i-Z_i\mut0-Z_i\ri (\mut1-\mut0)\\
    Z_i\ri Y_i -Z_i\ri\mut0-Z_i\ri^2(\mut1-\mut0)

\end{pmatrix}=\begin{pmatrix} 0\\0\\0\\0\end{pmatrix}
\end{equation}
